Assessment of simulation procedures and robust methods for Mendelian randomization analysis

### 2022-11-10 presentation/robustMR.qmd
update presentation

:::: {.columns}

::: {.column width="50%"}
* from the 284 SNPs only one was associated with the FTO gene
:::

::: {.column width="50%"}
![](../img/ensembl_logo.png)

:::

::::

@@

* se, pval and study size depend on each other, within each study
* selection of IVs based on exposure study (BMI in our case)
* Summary statistics on genetic associations with BMI based on results of the GIANT + GERA studies [@Hoffmann_2018] with $N=100,418 + 234,069 =334487$  and $SNPs=289$

@@

1. Calculate more accurates SEs using the betas and Pvals
2. Calculate the approximate SEs for decreasing samples sizes
3. Calculate the probability values considering the estimated betas and the corresponding SEs
4. Exclude the IV if Pval > genome-wide significance threshold  (5 × 10−8) **NEW 5x10^(-6))**
5. Calculate cumulative explained variances
6. Represent figure similar to Fig1 in Chatterjee’s paper @QiChatterjee_2021

@@

* The variance in BMI explained by the IVs is calculated with the following formula: $explained\ variance = 2f(1-f)\beta^2$, where $f$ denotes the allele frequency and $\beta$ is the additive genetic effect. 
* The explained variance in liability to GSD and GBC is calculated as described by @So_2011
  + Prevalence of disease $K$ must be set:
      - [Prevalence GSD](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3343155/#:~:text=Gallstones%20are%20common.,develop%20stones%20at%20some%20time.&text=The%20majority%20will%20not%20develop,cholecystitis,%20cholangitis,%20or%20pancreatitis)
      - [Prevalence GBC](https://www.wcrf.org/cancer-trends/gallbladder-cancer-statistics/)

@@

* Estimate parameters for the simulation (pi1, varx, pi2, varu, muy, vary, pi3, pi4, delta0) based on real data
* Start with $N=10.000$ and $N=300.000$

@@

>>>>

::: columns
::: {.column width="50%"}
-   from the 284 SNPs only one was associated with the FTO gene
:::

::: {.column width="50%"}
![](../img/ensembl_logo.png)
:::
:::

@@

-   se, pval and study size depend on each other, within each study
-   selection of IVs based on exposure study (BMI in our case)
-   Summary statistics on genetic associations with BMI based on results of the GIANT + GERA studies [@Hoffmann_2018] with $N=100,418 + 234,069 =334487$ and $SNPs=289$

@@

1.  Calculate more accurates SEs using the betas and Pvals
2.  Calculate the approximate SEs for decreasing samples sizes
3.  Calculate the probability values considering the estimated betas and the corresponding SEs
4.  Exclude the IV if Pval \> genome-wide significance threshold (5 × 10−8) **NEW 5x10\^(-6))**
5.  Calculate cumulative explained variances
6.  Represent figure similar to Fig1 in Chatterjee's paper @QiChatterjee_2021

@@

-   The variance in BMI explained by the IVs is calculated with the following formula: $explained\ variance = 2f(1-f)\beta^2$, where $f$ denotes the allele frequency and $\beta$ is the additive genetic effect.
-   The explained variance in liability to GSD and GBC is calculated as described by @So_2011
    -   Prevalence of disease $K$ must be set:
        -   [Prevalence GSD](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3343155/#:~:text=Gallstones%20are%20common.,develop%20stones%20at%20some%20time.&text=The%20majority%20will%20not%20develop,cholecystitis,%20cholangitis,%20or%20pancreatitis)
        -   [Prevalence GBC](https://www.wcrf.org/cancer-trends/gallbladder-cancer-statistics/)

@@

-   Estimate parameters for the simulation (pi1, varx, pi2, varu, muy, vary, pi3, pi4, delta0) based on real data
-   Start with $N=10.000$ and $N=300.000$

@@

## Meeting Summary I from 2022-11-09

### Update previous work

-   Harmonize 2-sample MR studies 
-   Explained Variance calculation:
    -   For BMI
        -   Use updated formula that takes sample size into account
    -   For GSD
        -   Update prevalence $K$ in accordance to epidemiological data from Michael Leitzmann and Frank Lammert
    -   For GBC
        -   Update prevalence $K=0.05$ from [Cancer Today](https://gco.iarc.fr/today/home)
-   Update Figure 01
-   Highlight FTO variant and update Figure 02

## Meeting Summary II from 2022-11-09

### Outlook

-   Calculate MR estimate of causal effect with different MR methods using:
    -   all significant IVs [adapted $pval-threshold=5x10^{-6}$]
    -   only FTO variant (get Wald estimate)

-   Simulation study: 
    -   Start with sample size of 50,000 and ~9 IVs (depending on how many remain after harmonizing datasets)
    -   Simulate for all significant IVs $\beta_{X_{sim}}$ from a normal distribution $\beta_{X_{sim}} \sim N(\beta_{\hat X_{IV}}, \hat se_{X_N})$ and $\beta_{X_{sim}}$ from a normal distribution $\beta_{Y_{sim}} \sim N(\beta_{\hat Y_{IV}}, \hat se_{Y_N})$; start with $n_{sim}=100$
    -   Simulate $maf$ from a Binomial distribution with $maf_{X_{sim}} \sim Bin(maf, 2N)$ 
        - This I did not fully understand: We do not want to count the successes, but simulate the $maf$. Should we draw it from a normal distribution too?
